Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-21 main.tex
def of clean

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
We write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{variable renaming} is a substitution $\sigma$ with $\sigma(x)\in\VAR$ for all variables $x$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
structural induction on $A$:
  \begin{itemize}
    \item Trivial if $A$ is an atom.
    \item The induction step uses the following fact:
	  if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
          module of $\gG$ if and only if it is a module of $\gG[X]$.
  \end{itemize}
  \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
\juihsuan{I'm still looking for a simple proof}
\footnote{Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then $\graphof A
= \single x \ \single p(x) \ \single x \ \single q(x)$, the scope of each
$x$-binder contains all the vertices, in particular, the two $x$-binders.}

@@

A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\substof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\substof\phi$, also called the \bfit{substitution induced by $\phi$} is a variable renaming such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\substof\phi(\labelof v)$. Note that this entails that
$\phi$ maps binders to binders and literals to literals.
We say that a
fograph homomorphism $\phi\colon\gG \rightarrow \gH$ is
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binders $b$ in
$\gG$, the vertex $\phi(b)$ is an existential binder in $\gH$.

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{A}{
        \vlde{\set{\acD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}

@@

    be lifted.

>>>>

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\rsubstof}[1]{\rho_{\!#1}}
\newcommand{\dsubstof}[1]{\delta_{\!#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}
\newcommand{\labelof}[1]{\ell(#1)}
\newcommand{\form}[1]{\mathrm{fm}(#1)}
\newcommand{\formtree}[1]{\mathcal{F}(#1)}

\renewcommand{\phi}{\varphi}

@@

A \bfit{substitution} is a function $\sigma\colon\VAR\to\TERM$ that is
the identity almost everywhere. We denote substitutions as
$\sigma=\sublist{\subst{x_1}{t_1},\ldots,\subst{x_n}{t_n}}$, where
$\sigma(x_i)=t_i$ for $i=1..n$ and $\sigma(x)=x$ for all
$x\notin\set{x_1,\ldots,x_n}$. 
We write $A\sigma$ for the formula
obtained from $A$ by applying $\sigma$, i.e., by simultaneously
replacing all occurrences of $x_i$ by $t_i$.
A \bfit{variable renaming} is a substitution $\rho$ with $\rho(x)\in\VAR$ for all variables $x$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. p(x)) \vlor (\forall x. q(x))$, then
$\graphof A = \single x \ \single p(x) \ \single x \ \single q(x)$,
the scope of each $x$-binder contains all the vertices, in particular,
the two $x$-binders. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in $\graphof{(\exists x. p(x)) \vlor (\exists x. q(x))}$,
both $x$-binders are legal, as they are not in each others
scope.\todo{draw the picture}.  For this reason, we call a formula
\bfit{clean} if it does not contain subformulas of the form $(\forall
x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$, and no
$x$-quantified formula occurs as subformula of another $x$-quantified
formula. Then we have:

\begin{lemma}
  If $A$ is clean iff\/ $\graphof A$ is a fograph.
\end{lemma}

\begin{proof}
  Induction on $A$, using  Theorem~\ref{thm:cograph}.
\end{proof}

@@

A \bfit{fograph homomorphism} $\phi=\tuple{\phi,\rsubstof\phi}$ is a
pair where $\phi\colon\gG\to\gH$ is a graph homomorphism between the
underlying graphs, and $\rsubstof\phi$, also called the \bfit{substitution induced by $\phi$} is a variable renaming such that
for all $v\in\vG$ we have
$\labelof{\phi(v)}=\rsubstof\phi(\labelof v)$. Note that this entails that
$\phi$ maps binders to binders and literals to literals.
We say that a
fograph homomorphism $\phi\colon\gG \rightarrow \gH$ is
%%\bfit{label-preserving} if for all $v \in \vG, label(v) = label(f(v))$, and is
\bfit{existential-preserving} if for all existential binders $b$ in
$\gG$, the vertex $\phi(b)$ is an existential binder in $\gH$.

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{A}{
        \vlde{\set{\acD,\cfaD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}

@@

    be lifted.\todo{medial permutation!!!}
